Calendar-based graphics for visualizing people's daily schedules

### 2017-08-09 main.Rmd
minor tweaks

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualizing people's daily schedules"
  plain: "Calendar-based graphics for visualizing people's daily schedules"
abstract: >
  This paper describes a \code{frame\_calendar} function that organizes and displays temporal data, collected on sub-daily resolution, into a calendar layout. Calendars are broadly used in society to display temporal information, and events. The \code{frame\_calendar} uses linear algebra on the date  variable to create the layout. It utilizes the grammar of graphics to create the plots inside each cell, and thus synchronises neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior in Melbourne, Australia, based on counts which are captured at hourly intervals by sensors scattered around the city. Faceting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to  events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the R package \pkg{sugrrants}.

@@

(ref:time-series-plot-cap) Time series plot showing the number of pedestrians in 2016 measured at 3 different sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The weekly patterns look distinctive across these 3 sensors. There is an eye-catching spike which occurred at the State Library, caused by the annual White Night event on 20th of February.

@@

(ref:facet-time-cap) Hourly pedestrian counts for 2016 faceted by sensors and days of the week using lines. It features at least two types of seasons---time of day and day of week---across all the sensors. The white night effect increasing pedestrian counts can be seen on Saturday in February at State Library.

@@

In contrast, calendar-based graphics unpacks the temporal variable, at different resolutions, to digest multiple seasonalities, and special events. There is some existing work in this area. For example, @VanWijkCluster1999 developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colours indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasonal variation such as school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in two R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are limited to colour-encoding graphics and are unable to use time scales smaller than a day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from pedestrian sensor data, will be neglected through daily aggregation. Additionally, if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

@@

(ref:month-diagram-cap) Illustration of the indexing layout for cells in a month, where $k$ is day of the week, $g$ is day of the month, $i,j$ indicates grid position.

@@

The algorithm has several optional parameters that modify the layout, direction of display, scales, white space and switching to polar coordinates. These are accessible to the user by the inputs to the function  \code{frame_calendar}:

```
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE, 
  nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed",
  width = 0.95, height = 0.95
)
```

@@

The \code{free_wday} scales each week day together. It can be useful to comparing trends across week days, allowing relative patterns for weekends vs week days to be examined. Similarly, the \code{free_mday} uses free scaling for any day within a given month.

@@

We also offer languages other than English for text labelling. Figure \ref{fig:chn} shows the same plot as Figure \ref{fig:boxplot} labelled using simplified Chinese characters.

@@

Plots can be layered. The comparison of sensors can be done by overlaying plot the values for each (Figure \ref{fig:overlay}). Difference between the pedestrian patterns for Flinders Street and Flagstaff train stations can be seen. Both Flinders Street and Flagstaff train stations exhibit strong  commuters patterns, however, Flinders Street has higher pedestrian counts during the weekends and public holidays. This suggests that Flagstaff Station has limited functionality on non-work days, but other activities occur around the Flinders Street Station. From Figure \ref{fig:overlay} it can be seen that the State Library has a similar temporal trend to Flinders Street Station on non-work days. The nighttime events, such as White Night and New Year's Eve, have barely affected the operation of Flagstaff Station but heavily affected the incoming and outgoing traffic to Flinders Street Station and the State Library.

@@

(ref:facet-cap) Line charts, embedded in the $6 \times 2$ monthly calendar, coloured and faceted by the 3 sensors. The variations of an individual sensor are emphasised, and the shapes can be compared across the cells and sensors.

@@

The \code{frame_calendar} function is not constrained to line plots. The full range of ggplot geoms is essentially available. Figure \ref{fig:scatterplot} shows a lag scatterplot at Flinders Street Station, where the current hourly count is assigned to the \code{x} argument and the lagged hourly count to the \code{y} argument. This figure is organized in the daily calendar layout. Figure \ref{fig:scatterplot} indicates two primary patterns, strong autocorrelation on weekends, and weaker autocorrelation on work days. At the higher counts, on week days, the next hour sees possibly substantial increase or decrease in counts, essentially revealing a bimodal distribution of consecutive counts, as supported by Figure \ref{fig:flinders-2016}.

@@

The focus is on the western calendar layout because most countries have adopted this format. However the language of labels would differ across countries, and so support was added for changing the label language simply.

>>>>

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualizing people's daily schedules"
  plain: "Calendar-based graphics for visualizing people's daily schedules"
abstract: >
  This paper describes a \code{frame\_calendar} function that organizes and displays temporal data, collected on sub-daily resolution, into a calendar layout. Calendars are broadly used in society to display temporal information, and events. The \code{frame\_calendar} uses linear algebra on the date  variable to create the layout. It utilizes the grammar of graphics to create the plots inside each cell, and thus synchronizes neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior in Melbourne, Australia, based on counts which are captured at hourly intervals by sensors scattered around the city. Faceting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to  events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the R package \pkg{sugrrants}.

@@

(ref:time-series-plot-cap) Time series plots showing the number of pedestrians in 2016 measured at 3 different sensors in the city of Melbourne. Colored by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The weekly patterns look distinctive across these 3 sensors. There is an eye-catching spike which occurred at the State Library, caused by the annual White Night event on 20th of February.

@@

(ref:facet-time-cap) Hourly pedestrian counts for 2016 faceted by sensors and days of the week using lines. It features at least two types of seasons---time of day and day of week---across all the sensors. The White Night effect on increasing pedestrian counts can be seen on Saturday in February at the State Library.

@@

In contrast, calendar-based graphics unpack the temporal variable, at different resolutions, to digest multiple seasonalities, and special events. There is some existing work in this area. For example, @VanWijkCluster1999 developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colors indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasonal variation such as school vacations, all of which have influence over the turn-outs in the office. The calendar-based heatmap was implemented in two R packages: \pkg{ggTimeSeries} [@R-ggTimeSeries] and \pkg{ggcal} [@R-ggcal]. However, these techniques are limited to color-encoding graphics and are unable to use time scales smaller than a day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from the pedestrian sensor data, will be neglected through daily aggregation. Additionally, if simply using colored blocks rather than curves, it may become perceptually difficult to estimate the shape positions and changes, although using curves comes with the cost of more display capacity [@cleveland1984graphical; @lam2007overview].

@@

(ref:month-diagram-cap) Illustration of the indexing layout for cells in a month, where $k$ is day of the week, $g$ is day of the month, $(i, j)$ indicates grid position.

@@

The algorithm has several optional parameters that modify the layout, direction of display, scales, plot size and switching to polar coordinates. These are accessible to the user by the inputs to the function  \code{frame_calendar}:

```
frame_calendar(data, x, y, date, calendar = "monthly", dir = "h", 
               sunday = FALSE, nrow = NULL, ncol = NULL, polar = FALSE, 
               scale = "fixed", width = 0.95, height = 0.95)
```

@@

The \code{free_wday} scales each week day together. It can be useful to comparing trends across week days, allowing relative patterns for weekends versus week days to be examined. Similarly, the \code{free_mday} uses free scaling for any day within a given month.

@@

Most countries have adopted this western calendar layout, while the languages used for week day and month would be different across countries. We also offer languages other than English for text labelling. Figure \ref{fig:chn} shows the same plot as Figure \ref{fig:boxplot} labelled using simplified Chinese characters.

@@

Plots can be layered. The comparison of sensors can be done by overlaying plot the values for each (Figure \ref{fig:overlay}). Differences between the pedestrian patterns for Flinders Street and Flagstaff train stations can be seen. Both Flinders Street and Flagstaff train stations exhibit strong  commuters patterns, however, Flinders Street has higher pedestrian counts during the weekends and public holidays. This suggests that Flagstaff Station has limited functionality on non-work days, but other activities occur around the Flinders Street Station. From Figure \ref{fig:overlay} it can be seen that the State Library has a similar temporal trend to Flinders Street Station on non-work days. The nighttime events, such as White Night and New Year's Eve, have barely affected the operation of Flagstaff Station but heavily affected the incoming and outgoing traffic to Flinders Street Station and the State Library.

@@

(ref:facet-cap) Line charts, embedded in the $6 \times 2$ monthly calendar, colored and faceted by the 3 sensors. The variations of an individual sensor are emphasised, and the shapes can be compared across the cells and sensors.

@@

The \code{frame_calendar} function is not constrained to line plots. The full range of plotting capabilities in \pkg{ggplot2} is essentially available. Figure \ref{fig:scatterplot} shows a lag scatterplot at Flinders Street Station, where the current hourly count is assigned to the \code{x} argument and the lagged hourly count to the \code{y} argument. This figure is organized in the daily calendar layout. Figure \ref{fig:scatterplot} indicates two primary patterns, strong autocorrelation on weekends, and weaker autocorrelation on work days. At the higher counts, on week days, the next hour sees possibly substantial increase or decrease in counts, essentially revealing a bimodal distribution of consecutive counts, as supported by Figure \ref{fig:flinders-2016}.

@@

The focus is on the western calendar layout, because most countries have adopted this format. However the language of labels would differ across countries, and so support was added for changing the label language simply.

### 2017-08-09 main.tex
minor tweaks

\Abstract{
This paper describes a \code{frame\_calendar} function that organizes
and displays temporal data, collected on sub-daily resolution, into a
calendar layout. Calendars are broadly used in society to display
temporal information, and events. The \code{frame\_calendar} uses linear
algebra on the date variable to create the layout. It utilizes the
grammar of graphics to create the plots inside each cell, and thus
synchronises neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior in Melbourne, Australia,
based on counts which are captured at hourly intervals by sensors
scattered around the city. Faceting by the usual features such as day
and month, was insufficient to examine the behavior. Making displays on
a monthly calendar format helps to understand pedestrian patterns
relative to events such as work days, weekends, holidays, and special
events. The layout algorithm has several format options and variations.
It is implemented in the R package \pkg{sugrrants}.
}

@@

\caption[Time series plot showing the number of
pedestrians in 2016 measured at 3 different sensors in the city of
Melbourne. Coloured by the sensors, small multiples of lines show that
the foot traffic varies from one sensor to another in terms of both time
and number. The weekly patterns look distinctive across these 3 sensors.
There is an eye-catching spike which occurred at the State Library,
caused by the annual White Night event on 20th of February.]{Time series plot showing the number of
pedestrians in 2016 measured at 3 different sensors in the city of
Melbourne. Coloured by the sensors, small multiples of lines show that
the foot traffic varies from one sensor to another in terms of both time
and number. The weekly patterns look distinctive across these 3 sensors.
There is an eye-catching spike which occurred at the State Library,
caused by the annual White Night event on 20th of February.}\label{fig:time-series-plot}
\end{figure}
\end{CodeChunk}

@@

\caption[Hourly pedestrian counts for 2016 faceted by
sensors and days of the week using lines. It features at least two types
of seasons---time of day and day of week---across all the sensors. The
white night effect increasing pedestrian counts can be seen on Saturday
in February at State Library.]{Hourly pedestrian counts for 2016 faceted by
sensors and days of the week using lines. It features at least two types
of seasons---time of day and day of week---across all the sensors. The
white night effect increasing pedestrian counts can be seen on Saturday
in February at State Library.}\label{fig:facet-time}
\end{figure}
\end{CodeChunk}

@@

In contrast, calendar-based graphics unpacks the temporal variable, at
different resolutions, to digest multiple seasonalities, and special
events. There is some existing work in this area. For example,
\citet{VanWijkCluster1999} developed a calendar view of the heatmap to
represent the number of employees in the work place over a year, where
colours indicate different clusters derived from the days. It contrasts
week days and weekends, highlights public holidays, and presents other
known seasonal variation such as school vacations, all of which have
influence over the turn-outs in the office. The calendar-based heatmap
was implemented in two R packages: \pkg{ggTimeSeries}
\citep{R-ggTimeSeries} and \pkg{ggcal} \citep{R-ggcal}. However, these
techniques are limited to colour-encoding graphics and are unable to use
time scales smaller than a day. Time of day, which serves as one of the
most important aspects in explaining substantial variations arising from
pedestrian sensor data, will be neglected through daily aggregation.
Additionally, if simply using colored blocks rather than curves, it may
become perceptually difficult to estimate the shape positions and
changes, although using curves comes with the cost of more display
capacity \citep{cleveland1984graphical, lam2007overview}.

@@

\caption[Illustration of the indexing layout for cells in
a month, where \(k\) is day of the week, \(g\) is day of the month,
\(i,j\) indicates grid position.]{Illustration of the indexing layout for cells in
a month, where \(k\) is day of the week, \(g\) is day of the month,
\(i,j\) indicates grid position.}\label{fig:month-diagram}
\end{figure}
\end{CodeChunk}

@@

The algorithm has several optional parameters that modify the layout,
direction of display, scales, white space and switching to polar
coordinates. These are accessible to the user by the inputs to the
function \code{frame_calendar}:

\begin{verbatim}
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE, 
  nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed",
  width = 0.95, height = 0.95
)
\end{verbatim}

@@

\caption[Line charts, embedded in the \(6 \times 2\) monthly
calendar, coloured and faceted by the 3 sensors. The variations of an
individual sensor are emphasised, and the shapes can be compared across
the cells and sensors.]{Line charts, embedded in the \(6 \times 2\) monthly
calendar, coloured and faceted by the 3 sensors. The variations of an
individual sensor are emphasised, and the shapes can be compared across
the cells and sensors.}\label{fig:facet}
\end{figure}
\end{CodeChunk}

>>>>

\Abstract{
This paper describes a \code{frame\_calendar} function that organizes
and displays temporal data, collected on sub-daily resolution, into a
calendar layout. Calendars are broadly used in society to display
temporal information, and events. The \code{frame\_calendar} uses linear
algebra on the date variable to create the layout. It utilizes the
grammar of graphics to create the plots inside each cell, and thus
synchronizes neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior in Melbourne, Australia,
based on counts which are captured at hourly intervals by sensors
scattered around the city. Faceting by the usual features such as day
and month, was insufficient to examine the behavior. Making displays on
a monthly calendar format helps to understand pedestrian patterns
relative to events such as work days, weekends, holidays, and special
events. The layout algorithm has several format options and variations.
It is implemented in the R package \pkg{sugrrants}.
}

@@

\caption[Time series plots showing the number of
pedestrians in 2016 measured at 3 different sensors in the city of
Melbourne. Colored by the sensors, small multiples of lines show that
the foot traffic varies from one sensor to another in terms of both time
and number. The weekly patterns look distinctive across these 3 sensors.
There is an eye-catching spike which occurred at the State Library,
caused by the annual White Night event on 20th of February.]{Time series plots showing the number of
pedestrians in 2016 measured at 3 different sensors in the city of
Melbourne. Colored by the sensors, small multiples of lines show that
the foot traffic varies from one sensor to another in terms of both time
and number. The weekly patterns look distinctive across these 3 sensors.
There is an eye-catching spike which occurred at the State Library,
caused by the annual White Night event on 20th of February.}\label{fig:time-series-plot}
\end{figure}
\end{CodeChunk}

@@

\caption[Hourly pedestrian counts for 2016 faceted by
sensors and days of the week using lines. It features at least two types
of seasons---time of day and day of week---across all the sensors. The
White Night effect on increasing pedestrian counts can be seen on
Saturday in February at the State Library.]{Hourly pedestrian counts for 2016 faceted by
sensors and days of the week using lines. It features at least two types
of seasons---time of day and day of week---across all the sensors. The
White Night effect on increasing pedestrian counts can be seen on
Saturday in February at the State Library.}\label{fig:facet-time}
\end{figure}
\end{CodeChunk}

@@

In contrast, calendar-based graphics unpack the temporal variable, at
different resolutions, to digest multiple seasonalities, and special
events. There is some existing work in this area. For example,
\citet{VanWijkCluster1999} developed a calendar view of the heatmap to
represent the number of employees in the work place over a year, where
colors indicate different clusters derived from the days. It contrasts
week days and weekends, highlights public holidays, and presents other
known seasonal variation such as school vacations, all of which have
influence over the turn-outs in the office. The calendar-based heatmap
was implemented in two R packages: \pkg{ggTimeSeries}
\citep{R-ggTimeSeries} and \pkg{ggcal} \citep{R-ggcal}. However, these
techniques are limited to color-encoding graphics and are unable to use
time scales smaller than a day. Time of day, which serves as one of the
most important aspects in explaining substantial variations arising from
the pedestrian sensor data, will be neglected through daily aggregation.
Additionally, if simply using colored blocks rather than curves, it may
become perceptually difficult to estimate the shape positions and
changes, although using curves comes with the cost of more display
capacity \citep{cleveland1984graphical, lam2007overview}.

@@

\caption[Illustration of the indexing layout for cells in
a month, where \(k\) is day of the week, \(g\) is day of the month,
\((i, j)\) indicates grid position.]{Illustration of the indexing layout for cells in
a month, where \(k\) is day of the week, \(g\) is day of the month,
\((i, j)\) indicates grid position.}\label{fig:month-diagram}
\end{figure}
\end{CodeChunk}

@@

The algorithm has several optional parameters that modify the layout,
direction of display, scales, plot size and switching to polar
coordinates. These are accessible to the user by the inputs to the
function \code{frame_calendar}:

\begin{verbatim}
frame_calendar(data, x, y, date, calendar = "monthly", dir = "h", 
               sunday = FALSE, nrow = NULL, ncol = NULL, polar = FALSE, 
               scale = "fixed", width = 0.95, height = 0.95)
\end{verbatim}

@@

\caption[Line charts, embedded in the \(6 \times 2\) monthly
calendar, colored and faceted by the 3 sensors. The variations of an
individual sensor are emphasised, and the shapes can be compared across
the cells and sensors.]{Line charts, embedded in the \(6 \times 2\) monthly
calendar, colored and faceted by the 3 sensors. The variations of an
individual sensor are emphasised, and the shapes can be compared across
the cells and sensors.}\label{fig:facet}
\end{figure}
\end{CodeChunk}
